---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# `rticulate`: Ultrasound Tongue Imaging in R

<!-- badges: start -->
<!-- badges: end -->

This is the repository of the R package `rticulate`. This package provides two functions for importing UTI data from Articulate Assistant Advanced and plotting UTI data from any source.

## Installation

To install the package, run the following (you need the package [remotes](https://remotes.r-lib.org)).

```{r install-latest, eval=FALSE}
remotes::install_github(
  "stefanocoretta/rticulate",
  ref = remotes::github_release(), # this pulls the latest release
  build_vignettes = TRUE
)
```

If you want to install the development version, run this instead.

```{r install-dev, eval=FALSE}
remotes::install_github(
  "stefanocoretta/rticulate",
  build_vignettes = TRUE
)
```

## Usage

To learn the basic use of the package, run `vignette("tongue-imaging", package = "rticulate")` after the installation.
