---
title: "Using generalised additive models (GAM) with polar coordinates for assessing tongue contours"
author: "Stefano Coretta"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tidyverse)
theme_set(theme_bw())
library(rticulate)
```

# Read data

The package `rticulate` comes with a dataset containg spline data from two speakers of Italian.

```{r load-data}
library(rticulate)
data(tongue)
tongue
```

# Fit a polar GAM

The spline data is in polar coordinates. The function `polar_gam()` converts the coordinates to polar and fits a GAM to the data.
The polar coordinates are calculated based on the origin of the probe, which is estimated if `origin = NULL` using the fan lines specified with the argument `fan_lines` (the defaults are `c(10, 25)`).

```{r polar-place}
tongue_it01 <- filter(tongue, speaker == "it01")

polar_place <- polar_gam(
    Y ~
        s(X, by = c2_place) +
        s(X, word, bs = "fs"),
    data = tongue_it01
)
summary(polar_place)
```

The output model is in polar coordinates but it contains the origin coordinates so that plotting can be done in cartesian coordinates.

We can now plot the results from the model with `plot_polar_smooths()`.

```{r plot-smooths}
plot_polar_smooths(
    polar_place,
    X,
    c2_place
) +
    theme(legend.position = "top")
```

